Constraining gravitational wave amplitude birefringence with GWTC-3

### 2023-02-28 src/tex/ms.tex
GW200129 exposition

\subsubsection{GW170818}
GW170818 gave the largest tension with $\kappa=0$.
Figure \ref{fig:corner_GW170818} shows the birefringence effect allows a region of $\kappa>0$ for
 larger distance and a smaller inclination angles than would be allowed in the absence of birefringence.
We can make sense of this result because a smaller inclination angle will make the amplitude ratio between the two circular polarizations smaller.
Then the birefringence effect will increase the amplitude ratio during the propagation.
The larger distance is for compensating the combined effect of a nonzero $\kappa$ and a smaller inclination angle.
We can see this effect in other events as well, but not as strongly as in GW170818.

@@

\subsubsection{GW200129}
\label{sec:GW200129}
Past research suggests a potential glitch in GW200129 Virgo data. \citep{GW200129_glitch}
We perform three sets of \ac{PE} to check if the glitch would affect the population posterior of $\kappa$.
We pick only two out of three detector data in each set to perform \ac{PE}. (i.e. LIGO-Hanford and LIGO-Livingston for the first set, LIGO-Hanford and Virgo for the second set, and LIGO-Livingston and Virgo for the third set.)
The two sets of \ac{PE} results with Virgo data return a $\kappa$ posterior that strongly suggests a negative $\kappa$.
Further investigation is needed to understand the glitch.
% Thus, we did not include this event in our collective analysis.

Figure~\ref{fig:corner_GW200129} shows the posteriors of the parameters of GW200129 are strongly affected by the Virgo data.

\begin{figure}
    \script{corner_GW200129.py}
    \includegraphics[width=\columnwidth]{figures/corner_GW200129.pdf}
    \caption{
        The posterior of $\kappa$, luminosity distance $d_L$ and $\cos{\iota}$ for GW200129.
        Colors in the plot are the \ac{PE} result with GR done by LVK without cosmological reweighing \citep{GWTC-2.1, GWTC-3, GWTC-2.1_dataset, GWTC-3_dataset} and the \ac{PE} result done by us with and without Virgo data.
        The 2D contours correspond to the $39.35\%$ ($1\sigma$) and $90\%$ credible levels.
        This plot shows that Virgo data may contain a glitch, which affects the \ac{PE} result.
    }
    \label{fig:corner_GW200129}
\end{figure}

>>>>

\subsubsection{GW170818}
GW170818 gave the largest tension with $\kappa=0$, when judged by $\mu_i/\sigma_i$ in Fig.~\ref{fig:violin_kappa}.
Figure \ref{fig:corner_GW170818} shows the birefringence effect allows a region of $\kappa>0$ for
 larger distance and a smaller inclination angles than would be allowed in the absence of birefringence.
We can make sense of this result because a smaller inclination angle will make the amplitude ratio between the two circular polarizations smaller.
Then the birefringence effect will increase the amplitude ratio during the propagation.
The larger distance is for compensating the combined effect of a nonzero $\kappa$ and a smaller inclination angle.
We can see this effect in other events as well, but not as strongly as in GW170818.

@@

\subsubsection{GW200129}
\label{sec:GW200129}

GW200129 is the event with the second largest $|\mu_i/\sigma_i|$ in Fig.~\ref{fig:violin_kappa}.
However, data for this event were affected by a non-Gaussian noise disturbance (glitch) in the Virgo instrument, which was subtracted in the publicly-available data used for parameter estimation \cite{Davis:2022ird}.
Since previous work suggests the degree of glitch subtraction affects the inference for this event \citep{GW200129_glitch}, we consider whether the apparent preference for $\kappa < 0$ could also be tied to the instrumental artifact.

To this end, we perform three additional \ac{PE} runs for GW200129, considering only two detectors at a time: LIGO Hanford and Virgo (HL), LIGO Livingston and Virgo (LV), and LIGO Hanford and LIGO Livingston (HL).
If the preference for $\kappa < 0$ is tied to the glitch in Virgo, we expect it to disappear in the HL run, which excludes Virgo data.

This is indeed the case, as we show in Fig.~\ref{fig:corner_GW200129}: all runs including Virgo lean towards $\kappa < 0$ (solid curves in color), whereas the LIGO-only run is fully consistent with $\kappa = 0$ (dashed black).
While this is not conclusive proof that the glitch itself is driving the result, it does indicate that the Virgo data play a key role in the inference of $\kappa$.
Since more work would be needed to understand the effect of the glitch, this motivates us to consider the effect of excluding this event from the collective analyses above.

\begin{figure}
    \script{corner_GW200129.py}
    \includegraphics[width=\columnwidth]{figures/corner_GW200129.pdf}
    \caption{
        GW200129 posterior on $\kappa$, luminosity distance $d_L$ and inclination $\cos{\iota}$, including different sets of detectors in the analysis per the legend.
        The main run with all three detectors (HLV, filled blue), shows a preference for $\kappa < 0$, as in Fig.~\ref{fig:violin_kappa}; this preference is more pronounced for two-detector runs that include Virgo (HV and LV, orange and green); however, it disappears if we remove Virgo (HL, dashed black).
        The 2D contours correspond to the $90\%$ credible level.
    }
    \label{fig:corner_GW200129}
\end{figure}
